---
title: "Untitled"
format: html
editor: visual
---
#1. Packages
```{r}
#| message: false
library(readr)
library(tidyverse)
library(seastests)
library(tsoutliers)
library(tseries)
library(corrplot)
library(fBasics)
library(EnvStats)
```

# 2.Data
```{r}
wd<-getwd()
setwd(wd)

can_data<-read_rds("can_data.rds")

```



#3. Diagnostics

## Data type
```{r}
str(can_data)
```


## Seasonality
```{r}
# Function
check_seasonality<-function(data){
  
  seasonal_data<-c()
  
  # We collect numeric data names
  num_columns<-names(data)[sapply(data,is.numeric)]
  
  # For each column of our data we apply this porcedure to test seasonality
  for(col in num_columns){
    
    # We select a single vector
    serie<-data[[col]]
    
    # We transform it into a time serie
    ts_serie<-ts(serie,freq=12)
    
    #we apply the test
    test_result<-combined_test(ts_serie)
    
    # Store the test result in a vector if there is seasonality
    if(test_result$stat== TRUE){
      seasonal_data<-c(seasonal_data,col)
  } 
  if (length(seasonal_data)==0){
   message("No signs of seasonality detected ") 
  }
  else{
    warning("Seasonality detected in:" )
    print(seasonal_data)
  }
  
    return(seasonal_data)
  }
}
```

```{r}
check_seasonality(can_data)
```


## Stationarity

```{r}
check_stationarity<-function(data){
  
  unitroot_data<-c()
  
  # We collect numeric data names
  num_columns<-names(data)[sapply(data,is.numeric)]
  
  # For each column of our data we apply this porcedure to test stationarity
  for(col in num_columns){
    
    # We select a single vector
    serie<-data[[col]]
    
    # We transform it into a time serie
    ts_serie<-ts(serie,freq=12)
    
    #we apply the test
    test_result<-adf.test(ts_serie)
    
    # Store the test result in a vector if there is seasonality
    if(test_result$p.value>0.05){
      unitroot_data<-c(unitroot_data,col)
  } 
  if (length(unitroot_data)==0){
   message("No signs of unit root detected ") 
  }
  else{
    warning("Unit root  detected in:" )
    print(unitroot_data)
  }
  
    return(unitroot_data)
  }
}
```

```{r}
check_stationarity(can_data)
```

#4. Housing variable (Y)

## Graph
```{r}
house<-ts(can_data$hstart_CAN_new,start=c(1990,1),frequency=12)

plot.ts(house,col="gray",main="Housing start monthly growth rate in Canada ")
abline(h=mean(house,na.rm=TRUE),col="salmon",lty=2)
```

#5. Exploratory data analysis

## Outliers of Y

```{r}
outlier<-tso(house)
plot(outlier)
show(outlier)
```
**Comments:** There appears to be a negative outlier in April 1990, classified as a TC (temporary change), indicating an initial shock followed by a return to normal. Possible causes include the recessionary cycle affecting the country during this period, partly driven by rising interest rates aimed at containing inflation.

### boxplot
```{r}
boxplot(house)
```

```{r}
rosner_test<-rosnerTest(house,k=10)
rosner_test$all.stats
```

**Comments:** Despite the potential outliers identified in the boxplot it looks like there are not statisicallly significant enough to be considered as such .

## Descriptive statistics of Y
```{r}
basicStats(house)
```


## Variables correlation:

### Correlation between housing and other variables

```{r}
thresold_cor <- 0.1

cor_vec <- cor(can_data |> select(-c(Date)), use = "complete.obs",method=c("spearman"))[, "hstart_CAN_new"]

cor_vec_filter <- cor_vec[abs(cor_vec) >= thresold_cor]

pal <- colorRampPalette(c("blue", "white", "red"))
cols <- pal(100)[as.numeric(cut(cor_vec_filter, breaks = 100))]

par(mar = c(45, 12, 4, 2))

barplot(cor_vec_filter,las=2,col = cols,
        main = paste0("Explanatory variables with correlation ≥ ", thresold_cor, "with hstart_CAN_new"),
        ylab = "Correlation", cex.names = 0.45,cex.main=0.9)

```
### Correlation between explanatory variables
```{r}
thresold <- 0.9

cor_mat <- can_data |>
  select(-c(Date,hstart_CAN_new)) |>
  cor(use = "complete.obs",method=c("spearman"))
diag(cor_mat) <- NA

vars_strong_co <- apply(cor_mat, 1, function(x) any(abs(x) > thresold, na.rm = TRUE))
cor_mat_filter <- cor_mat[vars_strong_co, vars_strong_co]

corrplot(cor_mat_filter, method = "circle", type = "upper", diag = FALSE, tl.cex = 0.7)

```






